 
\documentstyle[11pt]{article} 
\topmargin -.6in 
\textheight 8.7in 
\oddsidemargin -.0in 
\textwidth 6.5in 
\title{The Analysis of Starlight: Lab Projects} 
\date{Fall 1997} 
%\renewcommand{\baselinestretch}{1.2} 
\begin{document} 
\setcounter{page}{1} 
\setcounter{equation}{0} 
\pagestyle{empty} 
\parindent 0pt 
\parskip 8pt 
%\pagestyle{myheadings} 
\markright{{\bf LAB B: Measuring the Wavelength of Light} \
\hrulefill \ } 
\def\arcsec{\ifmmode {^{\scriptscriptstyle\prime\prime}}
          \else $^{\scriptscriptstyle\prime\prime}$\fi}
\def\arcmin{\ifmmode {^{\scriptscriptstyle\prime}}
          \else $^{\scriptscriptstyle\prime}$\fi}
\def\deg{\ifmmode^\circ\else$^\circ$\fi}


   
 
\noindent 
%\vspace{0.15in} 
{\bf Observational Astronomy      \hfill  V85.0013}\\ 
{\bf R1-S} 
 
\bigskip 
 
\bigskip 
 
\noindent 
{\hfill \Large {\bf Review Sheet 1} \hfill} 
 
 
\bigskip 

\begin{enumerate} 

\item 
In which constellation is the star Polaris (find it in Atlas Map 1).

\item
In which constellation is the asterism the Big Dipper (find it in Atlas Map 1).

\item
Most people know the signs of the Zodiac which correspond to the
constellations along the Zodiac in the sky: Aries, Taurus, ... What
three come  next.

\item
The greek alphabet goes: $\alpha$ alpha, $\beta$ beta, $\gamma$ gamma,
$\delta$ delta ... what three come next.

\item
What is the approximate angular width (in degrees) of a pinky at arms length.

\item
How many (10\deg) fists at arms length end-to-end, reach from the
horizon due north to the horizon due east.

\item
How many (10\deg) fists at arms length end-to-end, reach from the
horizon to the zenith.

\item
The angular diameter of the Moon is about 0.5\deg. How many Moons are
needed, edge-to-edge to reach from the horizon to the zenith.

\item
Your pinky at arms length can cover up the Moon: true or false.

\item
What is the altitude and azimuth of a star due east, half way up the
sky from the horizon.

\item
What is the angle between the zenith and the position alt = 0\deg, az
= 45\deg.

\item
Read this precise angle: 20\deg\ 35\arcmin\ 40\arcsec.

\item
How many \arcsec\ are there in 1\arcmin.

\item
What is the approximate angular diameter of the Moon in \arcmin. 

\item
What is the approximate angular size of your pinky at arms length, in \arcmin.

\item
What property would a star labeled beta (followed by the constellation
name) normally have.

\item Which animals correspond to the following constellations: Aries,
Ursa Minor, Canis Major

\item 
The Field Guide Fig 4-2 shows a picture of  Ursa Major from the
1801 Bode atlas.   Using the Atlas (map 1) , find the proper name
of the star at the end of the tail.

\item 
The Field Guide, Fig 2-13 shows the 1603 Beyer atlas picture of Orion. From
the Atlas (map 3) find the proper names of the two bright stars at
Orion's shoulders.

\item 
The Field Guide, Fig 2-5 shows the Beyer atlas picture of Cassiopeia. From
the Atlas (map 1) find the proper name of the  bright star at
Cassiopeia's throat.

\end{enumerate} 


\end{document}










