\documentclass[12pt]{article}
\usepackage[letterpaper]{geometry}

\frenchspacing
\pagestyle{empty}

\begin{document}
\section*{Final Exam, 2023 Fall}
\noindent
Name:
\vspace{0.25in}

\noindent
If you worked with anyone else when you completed this Exam, list their names
here, and describe (briefly) the help you gave one another:
\vspace{1in}

\begin{enumerate}
\item When we set up the telescope outside to start the night, what are
  the two stars we point at first?
  \vspace{80pt}
\item If an equatorial telescope is tracking, how long does it take to
  complete one full revolution? 
\vspace{80pt}
\item Say you observe a star at Dec = 0 passing through a telescope's
  field of view in 4 minutes. What is the angular size of that field
  of view?
\vspace{80pt}
\item You go outside one night at midnight standard time in New York
  State, and see Merak and Dubhe pointing down vertically at
  Polaris. What month is it? 
\vspace{80pt}
\item What is the sidereal period of the Moon's orbit?
      What is the synodic period of the Moon's orbit?
      Describe the difference between the two.
\vspace{80pt}
\item What are Kepler's laws?
\vspace{160pt}
\item How old is the Sun and how long will it exist for?
\vspace{80pt}
\item Why does high tide not always occur when the Moon transits?
\vspace{80pt}
\item Describe one way to determine the approximate mass of a binary
  star.
\vspace{80pt}
\item What are the four Galilean moons?  Very approximately, what are
  their periods? (That is: hours? days? weeks? months? years?).
\vspace{80pt}
\item What are the three galaxies other than the Milky Way visible with
  the naked eye? Which can you see from the Northern Hemisphere?
\vspace{80pt}
\end{enumerate}
\end{document}
