
\noindent
{Objectives:} To introduce some practical aspects of using an astronomical telescope.

\bigskip\noindent
{\bf INSIDE PREPARATIONS}

\input{dialing_setup.tex}

\clearpage
\noindent
{\bf OUTSIDE}

\bigskip
\noindent
{\bf 3. Setup}

\medskip\noindent
Set up the telescope according to the recipe given above, using as
reference star for RA that given below. The stars listed here are
appropriate for January or February lab times.

\medskip\begin{center}
NAME: \makebox[2cm]{Betelgeuse} RA: \makebox[2.5cm]{05h~55m}  Dec:
\makebox[2.5cm]{+07$^\circ$ 24$'$}
\end{center}

\bigskip\noindent
{\bf 4. Measuring Coordinates}

\medskip\noindent
Measure the coordinates of each of the stars in the table below. In
turn: locate the star by eye (e.g., using the sky maps in the field
guide after p 53); move the telescope so that the star is centered
first in the finder, then in the eyepiece; read off the star's
coordinates from the setting circles to within 5 min and 1\deg\ in RA
and Dec, respectively.

\begin{center}
\begin{tabular}{lcc} \hline \\ [-6pt]
\hspace{1cm}Name\hspace{1cm} &  \hspace{1cm} Your measured:  RA \hspace{1cm} & Dec \\ [6pt]
\hline
a) Betelguese  & &     \\ \hline
b) Castor  & &  \\ \hline
c) Algol  & &    \\ \hline
d) Sirius  & &   \\ \hline
  \end{tabular}
\end{center}

\bigskip\noindent
{\bf 5. Dialing objects}

\bigskip\noindent The table below lists the coordinates of four
objects. Dial them in on the setting circles and report what you see
at these locations. If you don't see anything obvious on the first
one, you probably have set up your telescope wrong!

\begin{center}
\begin{tabular}{lcc} \hline \\ [-6pt]
 \hspace{1.5cm}  RA \hspace{1cm} & Dec \hspace{1cm}& \hspace{3cm}Report\hspace{3cm} \\ [6pt]
\hline
b) 7h~39.29m  & $+5^\circ$ 13.3$'$ &     \\ \hline % Procyon
b) 3h~48.49m  & $+24^\circ$ 6.3$'$ &     \\ \hline % Pleiades
c) 5h~52.28m  & $+32^\circ$ 32.7$'$ &  \\ \hline % M37
d) 5h~35.14m   & $-5^\circ$ 25.9$'$ &    \\ \hline % Great Nebula in Orion
  \end{tabular}
\end{center}
